iSEE: Interactive SummarizedExperiment Explorer

### 2018-04-28 manuscript/main.tex
From Kevin: More URL links and conclusion. Added F1000 style.

\title{iSEE: Interactive SummarizedExperiment Explorer}
\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise. Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
\author[1,$\dagger$]{Kevin Rue-Albrecht}
\author[2,3,$\dagger$]{Federico Marini}
\author[4,$\dagger$]{Charlotte Soneson}
\author[5,$\dagger$,$\ast$]{Aaron T. L. Lun}

@@

\begin{abstract}

\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (\url{https://bioconductor.org/packages/iSEE}).\\
\textbf{Contact:} aaron.lun@cruk.cam.ac.uk\\
\textbf{Supplementary information:} Supplementary data are available online.

@@

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).

@@

\subsection*{Single-cell droplet-based RNA-seq}
% TODO: better title, or do we drop the allen section altogether (above)?
We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing dataset from 10X Genomics \citep{zheng2017massively}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
% embed the allen demo data set
\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}

\subsection*{Bulk RNA-seq from TCGA}
We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
% TODO: embed the TCGA demo data set
\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}

\subsection*{CyTOF}
Lastly, we also showcase the use of iSEE to vizualise and refine gating a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
% TODO: repurpose the conclusion if we showcase the examples in earlier sections
% TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest. Something like...
We expect the functionality and flexibility of the iSEE package to be essential when exploring and showcasing relevant features of any dataset which can be represented in the form of a SummarizedExperiment object, in such a way that a public instance of iSEE can ideally accompany any publication, e.g. reproducing the manuscript figures, but also allowing any user to foster hypothesis generation with efficient interactive and reproducible data analysis.

>>>>

\title{iSEE: Interactive SummarizedExperiment Explorer}
\author[1,$\dagger$]{Kevin Rue-Albrecht}
\author[2,3,$\dagger$]{Federico Marini}
\author[4,$\dagger$]{Charlotte Soneson}
\author[5,$\dagger$,$\ast$]{Aaron T. L. Lun}

@@

\begin{abstract}
Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.

@@

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).\\

@@

\subsection*{Single-cell droplet-based RNA-seq}
% TODO: better title, or do we drop the allen section altogether (above)?
We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing dataset from 10X Genomics \citep{zheng2017massively}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

% TODO: embed the allen demo data set
Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}

\subsection*{Bulk RNA-seq from TCGA}
We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

% TODO: embed the TCGA demo data set
Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}

\subsection*{CyTOF}
Lastly, we also showcase the use of iSEE to vizualise and refine a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
% TODO: repurpose the conclusion if we showcase the examples in earlier sections
% TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest. Something like...
We expect the functionalities and flexibility of the iSEE package to be essential for exploring and showcasing relevant features of any data set that can be represented in the form of a SummarizedExperiment object.
In particular, we expect public instances of iSEE to ideally accompany publications, as they enable authors to reproduce published figures and showcase salient aspects of their work using guided tours, while fostering hypothesis generation through transparent, interactive, reproducible, and efficient data analysis by the broader reader audience.
